### 2019-03-30 demo.tex
initial

%%%% Proceedings format for most of ACM conferences (with the exceptions listed below) and all ICPS volumes.
% \documentclass[sigconf]{acmart}
% \documentclass[acmtog]{acmart}
\documentclass[sigconf]{acmart}
\usepackage{graphicx}
\graphicspath{{imgs/}}
\usepackage{subfig}
\usepackage[english,ngerman,brazilian]{babel}
\settopmatter{printacmref=false}
\setcopyright{none}
\renewcommand\footnotetextcopyrightpermission[1]{}
\pagestyle{plain}

% defining the \BibTeX command - from Oren Patashnik's original BibTeX documentation.
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08emT\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

%
% end of the preamble, start of the body of the document source.
\begin{document}

%
% The "title" command has an optional parameter, allowing the author to define a "short title" to be used in page headers.
\title[Fronteiras da Transferência de Aprendizagem: uma revisão sistemática com enfoque meta-analítico]{Fronteiras da Transferência de Aprendizagem: \\
uma revisão sistemática com enfoque meta-analítico}

%
% The "author" command and its associated commands are used to define the authors and their affiliations.
% Of note is the shared affiliation of the first two authors, and the "authornote" and "authornotemark" commands
% used to denote shared contribution to the research.
\author{Fred Guth}
% \authornote{Both authors contributed equally to this research.}
\email{fredguth@fredguth.com}
\affiliation{%
  \institution{Departamento de Ciência de Computação, Universidade de Brasília}
  \postcode{70.910-900}
  \city{Brasília}
  \state{DF}
  \country{Brazil}
}

%
% By default, the full list of authors will be used in the page headers. Often, this list is too long, and will overlap
% other information printed in the page headers. This command allows the author to define a more concise list
% of authors' names for this purpose.
\renewcommand{\shortauthors}{Guth, F.}

%
% The abstract is a short summary of the work to be presented in the article.
\begin{abstract}
A clear and well-documented \LaTeX\ document is presented as an article formatted for publication by ACM in 
a conference proceedings or journal publication. Based on the ``acmart'' document class, this article presents
and explains many of the common variations, as well as many of the formatting elements
an author may use in the preparation of the documentation of their work.
\end{abstract}

%
% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
% Please copy and paste the code instead of the example below.
%
\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>10010147.10010257.10010258.10010262.10010277</concept_id>
  <concept_desc>Computing methodologies~Transfer learning</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Computing methodologies~Transfer learning}

\keywords{transferência de aprendizado, revisão sistemática, enfoque meta-analítico}

%
% A "teaser" image appears between the author and affiliation information and the body 
% of the document, and typically spans the page. 
% \begin{teaserfigure}
%   \includegraphics[width=\textwidth]{sampleteaser}
%   \caption{Seattle Mariners at Spring Training, 2010.}
%   \Description{Enjoying the baseball game from the third-base seats. Ichiro Suzuki preparing to bat.}
%   \label{fig:teaser}
% \end{teaserfigure}

%
% This command processes the author and affiliation and title information and builds
% the first part of the formatted document.
\maketitle

% 1 Introdução
% 1.1 Contribuições
% 1.2 Visão Geral e Organização do Artigo
% 1.3 Trabalhos Relacionados
% 2 Método de Revisão Sistemática com Enfoque Meta Analítico
% 2.1 o que é Enfoque Meta Analítico e por que adotá-lo
% 2.2 
% 3 Revisão da Literatura
% 3.1 O que é Transferência de Aprendizado e por que é estudada
% 3.1.1 Uma breve história da pesquisa na área
% 3.3 Os "Clássicos"
% 3.4 A "Fronteira"
% 4 Problemas Abertos e Implicações para Projetos de Pesquisa
% 5 Conclusão
% 6 Referências

\section{Introdução} 
  \subsection{Contribuições}
  \subsection{Visão Geral e Organização do Artigo}
  \subsection{Trabalhos Relacionados}
\section{Método de Revisão Sistemática}
  \subsection{O Enfoque Meta Analítico}
  \subsection{Análise de Co-citações}
  \subsection{Análise de Acoplamento Bibliográfico}
  \subsection{Análise Textual}
  \subsection{Sumarização}
\subsection{Revisão da Literatura}
  \subsection{Transferência de Aprendizagem}
  \subsection{Um breve histórico}
  \subsection{Os Clássicos}
  \subsection{A Fronteira}
\section{Problemas em Aberto}
\section{Conclusão}

In addition to specifying the {\it template style} to be used in formatting your work, there are a number of {\it template parameters} which modify some part of the applied template style. A complete list of these parameters can be found in the {\it \LaTeX\ User's Guide.}

Frequently-used parameters, or combinations of parameters, include:
\begin{itemize}
\item {\verb|anonymous,review|}: Suitable for a ``double-blind'' conference submission. Anonymizes the work and includes line numbers. Use with the \verb|\acmSubmissionID| command to print the submission's unique ID on each page of the work.
\item{\verb|authorversion|}: Produces a version of the work suitable for posting by the author.
\item{\verb|screen|}: Produces colored hyperlinks.
\end{itemize}

This document uses the following string as the first command in the source file: \verb|\documentclass[sigconf,screen]{acmart}|.

\section{Modifications}

Modifying the template --- including but not limited to: adjusting margins, typeface sizes, line spacing, paragraph and list definitions, and the use of the \verb|\vspace| command to manually adjust the vertical spacing between elements of your work --- is not allowed.

{\bf Your document will be returned to you for revision if modifications are discovered.}

\section{Typefaces}

The ``\verb|acmart|'' document class requires the use of the ``Libertine'' typeface family. Your \TeX\ installation should include this set of packages. Please do not substitute other typefaces. The ``\verb|lmodern|'' and ``\verb|ltimes|'' packages should not be used, as they will override the built-in typeface families.

\section{Title Information}

The title of your work should use capital letters appropriately - \url{https://capitalizemytitle.com/} has useful rules for capitalization. Use the {\verb|title|} command to define the title of your work. If your work has a subtitle, define it with the {\verb|subtitle|} command.
Do not insert line breaks in your title.

If your title is lengthy, you must define a short version to be used in the page headers, to prevent overlapping text. The \verb|title| command has a ``short title'' parameter:
\begin{verbatim}
  \title[short title]{full title}
\end{verbatim}

\section{Authors and Affiliations}

Each author must be defined separately for accurate metadata identification. Multiple authors may share one affiliation. Authors' names should not be abbreviated; use full first names wherever possible. Include authors' e-mail addresses whenever possible.

Grouping authors' names or e-mail addresses, or providing an ``e-mail alias,'' as shown below, is not acceptable:
\begin{verbatim}
  \author{Brooke Aster, David Mehldau}
  \email{dave,judy,steve@university.edu}
  \email{firstname.lastname@phillips.org}
\end{verbatim}

The \verb|authornote| and \verb|authornotemark| commands allow a note to apply to multiple authors --- for example, if the first two authors of an article contributed equally to the work.

If your author list is lengthy, you must define a shortened version of the list of authors to be used in the page headers, to prevent overlapping text. The following command should be placed just after the last \verb|\author{}| definition:
\begin{verbatim}
  \renewcommand{\shortauthors}{McCartney, et al.}
\end{verbatim}
Omitting this command will force the use of a concatenated list of all of the authors' names, which may result in overlapping text in the page headers.

The article template's documentation, available at  \url{https://www.acm.org/publications/proceedings-template}, has a complete explanation of these commands and tips for their effective use.

\section{Rights Information}

Authors of any work published by ACM will need to complete a rights form. Depending on the kind of work, and the rights management choice made by the author, this may be copyright transfer, permission, license, or an OA (open access) agreement.

Regardless of the rights management choice, the author will receive a copy of the completed rights form once it has been submitted. This form contains \LaTeX\ commands that must be copied into the source document. When the document source is compiled, these commands and their parameters add formatted text to several areas of the final document:
\begin{itemize}
\item the ``ACM Reference Format'' text on the first page.
\item the ``rights management'' text on the first page.
\item the conference information in the page header(s).
\end{itemize}

Rights information is unique to the work; if you are preparing several works for an event, make sure to use the correct set of commands with each of the works.

\section{CCS Concepts and User-Defined Keywords}

Two elements of the ``acmart'' document class provide powerful taxonomic tools for you to help readers find your work in an online search.

The ACM Computing Classification System --- \url{https://www.acm.org/publications/class-2012} --- is a set of classifiers and concepts that describe the computing discipline. Authors can select entries from this classification system, via \url{https://dl.acm.org/ccs/ccs.cfm}, and generate the commands to be included in the \LaTeX\ source.

User-defined keywords are a comma-separated list of words and phrases of the authors' choosing, providing a more flexible way of describing the research being presented.

CCS concepts and user-defined keywords are required for all short- and full-length articles, and optional for two-page abstracts.

\section{Sectioning Commands}

Your work should use standard \LaTeX\ sectioning commands: \verb|section|, \verb|subsection|, \verb|subsubsection|, and \verb|paragraph|. They should be numbered; do not remove the numbering from the commands.

Simulating a sectioning command by setting the first word or words of a paragraph in boldface or italicized text is {\bf not allowed.}

\section{Tables}

The ``\verb|acmart|'' document class includes the ``\verb|booktabs|'' package --- \url{https://ctan.org/pkg/booktabs} --- for preparing high-quality tables.

Table captions are placed {\it above} the table.

Because tables cannot be split across pages, the best placement for them is typically the top of the page nearest their initial cite.  To ensure this proper ``floating'' placement of tables, use the environment \textbf{table} to enclose the table's contents and the table caption.  The contents of the table itself must go in the \textbf{tabular} environment, to be aligned properly in rows and columns, with the desired horizontal and vertical rules.  Again, detailed instructions on \textbf{tabular} material are found in the \textit{\LaTeX\ User's Guide}.

Immediately following this sentence is the point at which Table~\ref{tab:freq} is included in the input file; compare the placement of the table here with the table in the printed output of this document.

\begin{table}
  \caption{Frequency of Special Characters}
  \label{tab:freq}
  \begin{tabular}{ccl}
    \toprule
    Non-English or Math&Frequency&Comments\\
    \midrule
    \O & 1 in 1,000& For Swedish names\\
    $\pi$ & 1 in 5& Common in math\\
    \$ & 4 in 5 & Used in business\\
    $\Psi^2_1$ & 1 in 40,000& Unexplained usage\\
  \bottomrule
\end{tabular}
\end{table}

To set a wider table, which takes up the whole width of the page's live area, use the environment \textbf{table*} to enclose the table's contents and the table caption.  As with a single-column table, this wide table will ``float'' to a location deemed more desirable. Immediately following this sentence is the point at which Table~\ref{tab:commands} is included in the input file; again, it is instructive to compare the placement of the table here with the table in the printed output of this document.

\begin{table*}
  \caption{Some Typical Commands}
  \label{tab:commands}
  \begin{tabular}{ccl}
    \toprule
    Command &A Number & Comments\\
    \midrule
    \texttt{{\char'134}author} & 100& Author \\
    \texttt{{\char'134}table}& 300 & For tables\\
    \texttt{{\char'134}table*}& 400& For wider tables\\
    \bottomrule
  \end{tabular}
\end{table*}

\section{Math Equations}
You may want to display math equations in three distinct styles:
inline, numbered or non-numbered display.  Each of
the three are discussed in the next sections.

\subsection{Inline (In-text) Equations}
A formula that appears in the running text is called an
inline or in-text formula.  It is produced by the
\textbf{math} environment, which can be
invoked with the usual \texttt{{\char'134}begin\,\ldots{\char'134}end}
construction or with the short form \texttt{\$\,\ldots\$}. You
can use any of the symbols and structures,
from $\alpha$ to $\omega$, available in
\LaTeX~\cite{Lamport:LaTeX}; this section will simply show a
few examples of in-text equations in context. Notice how
this equation:
\begin{math}
  \lim_{n\rightarrow \infty}x=0
\end{math},
set here in in-line math style, looks slightly different when
set in display style.  (See next section).

\subsection{Display Equations}
A numbered display equation---one set off by vertical space from the
text and centered horizontally---is produced by the \textbf{equation}
environment. An unnumbered display equation is produced by the
\textbf{displaymath} environment.

Again, in either environment, you can use any of the symbols
and structures available in \LaTeX\@; this section will just
give a couple of examples of display equations in context.
First, consider the equation, shown as an inline equation above:
\begin{equation}
  \lim_{n\rightarrow \infty}x=0
\end{equation}
Notice how it is formatted somewhat differently in
the \textbf{displaymath}
environment.  Now, we'll enter an unnumbered equation:
\begin{displaymath}
  \sum_{i=0}^{\infty} x + 1
\end{displaymath}
and follow it with another numbered equation:
\begin{equation}
  \sum_{i=0}^{\infty}x_i=\int_{0}^{\pi+2} f
\end{equation}
just to demonstrate \LaTeX's able handling of numbering.

\section{Figures}

The ``\verb|figure|'' environment should be used for figures. One or more images can be placed within a figure. If your figure contains third-party material, you must clearly identify it as such, as shown in the example below.
\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{sample-franklin}
  \caption{1907 Franklin Model D roadster. Photograph by Harris \& Ewing, Inc. [Public domain], via Wikimedia Commons. (\url{https://goo.gl/VLCRBB}).}
  \Description{The 1907 Franklin Model D roadster}
\end{figure}

Your figures should contain a caption which describes the figure to the reader. Figure captions go below the figure. Your figures should {\bf also} include a description suitable for screen readers, to assist the visually-challenged to better understand your work.

Figure captions are placed {\it below} the figure.

\subsection{The ``Teaser Figure''}

A ``teaser figure'' is an image, or set of images in one figure, that are placed after all author and affiliation information, and before the body of the article, spanning the page. If you wish to have such a figure in your article, place the command immediately before the \verb|\maketitle| command:
\begin{verbatim}
  \begin{teaserfigure}
    \includegraphics[width=\textwidth]{sampleteaser}
    \caption{figure caption}
    \Description{figure description}
  \end{teaserfigure}
\end{verbatim}

\section{Citations and Bibliographies}

The use of \BibTeX\ for the preparation and formatting of one's references is strongly recommended. Authors' names should be complete --- use full first names (``Donald E. Knuth'') not initials (``D. E. Knuth'') --- and the salient identifying features of a reference should be included: title, year, volume, number, pages, article DOI, etc.

The bibliography is included in your source document with these two commands, placed just before the \verb|\end{document}| command:
\begin{verbatim}
  \bibliographystyle{ACM-Reference-Format}
  \bibliography{bibfile}
\end{verbatim}
where ``\verb|bibfile|'' is the name, without the ``\verb|.bib|'' suffix, of the \BibTeX\ file.

Citations and references are numbered by default. A small number of ACM publications have citations and references formatted in the ``author year'' style; for these exceptions, please include this command in the {\bf preamble} (before ``\verb|\begin{document}|'') of your \LaTeX\ source: 
\begin{verbatim}
  \citestyle{acmauthoryear}
\end{verbatim}

Some examples.  A paginated journal article \cite{Abril07}, an enumerated journal article \cite{Cohen07}, a reference to an entire issue \cite{JCohen96}, a monograph (whole book) \cite{Kosiur01}, a monograph/whole book in a series (see 2a in spec. document)
\cite{Harel79}, a divisible-book such as an anthology or compilation \cite{Editor00} followed by the same example, however we only output the series if the volume number is given \cite{Editor00a} (so Editor00a's series should NOT be present since it has no vol. no.),
a chapter in a divisible book \cite{Spector90}, a chapter in a divisible book in a series \cite{Douglass98}, a multi-volume work as book \cite{Knuth97}, an article in a proceedings (of a conference, symposium, workshop for example) (paginated proceedings article) \cite{Andler79}, a proceedings article with all possible elements \cite{Smith10}, an example of an enumerated proceedings article \cite{VanGundy07}, an informally published work \cite{Harel78}, a doctoral dissertation \cite{Clarkson85}, a master's thesis: \cite{anisi03}, an online document / world wide web resource \cite{Thornburg01, Ablamowicz07, Poker06}, a video game (Case 1) \cite{Obama08} and (Case 2) \cite{Novak03} and \cite{Lee05} and (Case 3) a patent \cite{JoeScientist001}, work accepted for publication \cite{rous08}, 'YYYYb'-test for prolific author \cite{SaeediMEJ10} and \cite{SaeediJETC10}. Other cites might contain 'duplicate' DOI and URLs (some SIAM articles) \cite{Kirschmer:2010:AEI:1958016.1958018}. Boris / Barbara Beeton: multi-volume works as books \cite{MR781536} and \cite{MR781537}. A couple of citations with DOIs: \cite{2004:ITE:1009386.1010128,Kirschmer:2010:AEI:1958016.1958018}. Online citations: \cite{TUGInstmem, Thornburg01, CTANacmart}.

\section{Acknowledgments}

Identification of funding sources and other support, and thanks to individuals and groups that assisted in the research and the preparation of the work should be included in an acknowledgment section, which is placed just before the reference section in your document.

This section has a special environment:
\begin{verbatim}
  \begin{acks}
  ...
  \end{acks}
\end{verbatim}
so that the information contained therein can be more easily collected during the article metadata extraction phase, and to ensure consistency in the spelling of the section heading.

Authors should not prepare this section as a numbered or unnumbered {\verb|\section|}; please use the ``{\verb|acks|}'' environment.

\section{Appendices}

If your work needs an appendix, add it before the ``\verb|\end{document}|'' command at the conclusion of your source document.

Start the appendix with the ``\verb|appendix|'' command:
\begin{verbatim}
  \appendix
\end{verbatim}
and note that in the appendix, sections are lettered, not numbered. This document has two appendices, demonstrating the section and subsection identification method.

\section{SIGCHI Extended Abstracts}

The ``\verb|sigchi-a|'' template style (available only in \LaTeX\ and not in Word) produces a landscape-orientation formatted article, with a wide left margin. Three environments are available for use with the ``\verb|sigchi-a|'' template style, and produce formatted output in the margin:
\begin{itemize}
\item {\verb|sidebar|}:  Place formatted text in the margin.
\item {\verb|marginfigure|}: Place a figure in the margin.
\item {\verb|margintable|}: Place a table in the margin.
\end{itemize}

%
% The acknowledgments section is defined using the "acks" environment (and NOT an unnumbered section). This ensures
% the proper identification of the section in the article metadata, and the consistent spelling of the heading.
\begin{acks}
To Robert, for the bagels and explaining CMYK and color spaces.
\end{acks}

%
% The next two lines define the bibliography style to be used, and the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{sample-base}

% 
% If your work has an appendix, this is the place to put it.
\appendix

\section{Research Methods}

\subsection{Part One}

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi malesuada, quam in pulvinar varius, metus nunc fermentum urna, id sollicitudin purus odio sit amet enim. Aliquam ullamcorper eu ipsum vel mollis. Curabitur quis dictum nisl. Phasellus vel semper risus, et lacinia dolor. Integer ultricies commodo sem nec semper.

\subsection{Part Two}

Etiam commodo feugiat nisl pulvinar pellentesque. Etiam auctor sodales ligula, non varius nibh pulvinar semper. Suspendisse nec lectus non ipsum convallis congue hendrerit vitae sapien. Donec at laoreet eros. Vivamus non purus placerat, scelerisque diam eu, cursus ante. Etiam aliquam tortor auctor efficitur mattis.

\section{Online Resources}

Nam id fermentum dui. Suspendisse sagittis tortor a nulla mollis, in pulvinar ex pretium. Sed interdum orci quis metus euismod, et sagittis enim maximus. Vestibulum gravida massa ut felis suscipit congue. Quisque mattis elit a risus ultrices commodo venenatis eget dui. Etiam sagittis eleifend elementum.

Nam interdum magna at lectus dignissim, ac dignissim lorem rhoncus. Maecenas eu arcu ac neque placerat aliquam. Nunc pulvinar massa et mattis lacinia.

\bibliographystyle{ACM-Reference-Format}
\bibliography{references}
\end{document}

### 2019-03-30 paper.tex
initial

--- 2019-03-30 demo.tex
@@
 %%%% Proceedings format for most of ACM conferences (with the exceptions listed below) and all ICPS volumes.
-% \documentclass[sigconf]{acmart}
-% \documentclass[acmtog]{acmart}
 \documentclass[sigconf]{acmart}
@@
 \graphicspath{{imgs/}}
-\usepackage{subfig}
 \usepackage[english,ngerman,brazilian]{babel}
@@
 
-% defining the \BibTeX command - from Oren Patashnik's original BibTeX documentation.
 \def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08emT\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
 
-%
 % end of the preamble, start of the body of the document source.
@@
 uma revisão sistemática com enfoque meta-analítico}
-
 %
@@
 % of authors' names for this purpose.
-\renewcommand{\shortauthors}{Guth, F.}
+\renewcommand{\shortauthors}{Guth,.}
 
@@
 %
+
 \begin{CCSXML}
@@
  <concept>
-  <concept_id>10010147.10010257.10010258.10010262.10010277</concept_id>
-  <concept_desc>Computing methodologies~Transfer learning</concept_desc>
+  <concept_id>10010520.10010553.10010562</concept_id>
+  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
   <concept_significance>500</concept_significance>
+ </concept>
+ <concept>
+  <concept_id>10010520.10010575.10010755</concept_id>
+  <concept_desc>Computer systems organization~Redundancy</concept_desc>
+  <concept_significance>300</concept_significance>
+ </concept>
+ <concept>
+  <concept_id>10010520.10010553.10010554</concept_id>
+  <concept_desc>Computer systems organization~Robotics</concept_desc>
+  <concept_significance>100</concept_significance>
+ </concept>
+ <concept>
+  <concept_id>10003033.10003083.10003095</concept_id>
+  <concept_desc>Networks~Network reliability</concept_desc>
+  <concept_significance>100</concept_significance>
  </concept>
@@
 
-\ccsdesc[500]{Computing methodologies~Transfer learning}
+\ccsdesc[500]{Computer systems organization~Embedded systems}
+\ccsdesc[300]{Computer systems organization~Redundancy}
+\ccsdesc{Computer systems organization~Robotics}
+\ccsdesc[100]{Networks~Network reliability}
 
-\keywords{transferência de aprendizado, revisão sistemática, enfoque meta-analítico}
+%
+% Keywords. The author(s) should pick words that accurately describe the work being
+% presented. Separate the keywords with commas.
+\keywords{datasets, neural networks, gaze detection, text tagging}
 
@@
 % of the document, and typically spans the page. 
-% \begin{teaserfigure}
-%   \includegraphics[width=\textwidth]{sampleteaser}
-%   \caption{Seattle Mariners at Spring Training, 2010.}
-%   \Description{Enjoying the baseball game from the third-base seats. Ichiro Suzuki preparing to bat.}
-%   \label{fig:teaser}
-% \end{teaserfigure}
+\begin{teaserfigure}
+  \includegraphics[width=\textwidth]{sampleteaser}
+  \caption{Seattle Mariners at Spring Training, 2010.}
+  \Description{Enjoying the baseball game from the third-base seats. Ichiro Suzuki preparing to bat.}
+  \label{fig:teaser}
+\end{teaserfigure}
 
@@
 
-% 1 Introdução
-% 1.1 Contribuições
-% 1.2 Visão Geral e Organização do Artigo
-% 1.3 Trabalhos Relacionados
-% 2 Método de Revisão Sistemática com Enfoque Meta Analítico
-% 2.1 o que é Enfoque Meta Analítico e por que adotá-lo
-% 2.2 
-% 3 Revisão da Literatura
-% 3.1 O que é Transferência de Aprendizado e por que é estudada
-% 3.1.1 Uma breve história da pesquisa na área
-% 3.3 Os "Clássicos"
-% 3.4 A "Fronteira"
-% 4 Problemas Abertos e Implicações para Projetos de Pesquisa
-% 5 Conclusão
-% 6 Referências
+\section{Introduction}
+ACM's consolidated article template, introduced in 2017, provides a consistent \LaTeX\ style for use across ACM publications, and incorporates accessibility and metadata-extraction functionality necessary for future Digital Library endeavors.
+Numerous ACM and SIG-specific \LaTeX\ templates have been examined, and their unique features incorporated into this single new template.
 
-\section{Introdução} 
-  \subsection{Contribuições}
-  \subsection{Visão Geral e Organização do Artigo}
-  \subsection{Trabalhos Relacionados}
-\section{Método de Revisão Sistemática}
-  \subsection{O Enfoque Meta Analítico}
-  \subsection{Análise de Co-citações}
-  \subsection{Análise de Acoplamento Bibliográfico}
-  \subsection{Análise Textual}
-  \subsection{Sumarização}
-\subsection{Revisão da Literatura}
-  \subsection{Transferência de Aprendizagem}
-  \subsection{Um breve histórico}
-  \subsection{Os Clássicos}
-  \subsection{A Fronteira}
-\section{Problemas em Aberto}
-\section{Conclusão}
+If you are new to publishing with ACM, this document is a valuable guide to the process of preparing your work for publication.
+If you have published with ACM before, this document provides insight and instruction into more recent changes to the article template.
+
+The ``\verb|acmart|'' document class can be used to prepare articles for any ACM publication --- conference or journal, and for any stage of publication, from review to final ``camera-ready'' copy, to the author's own version, with {\it very} few changes to the source.
+
+\section{Template Overview}
+As noted in the introduction, the ``\verb|acmart|'' document class can be used to prepare many different kinds of documentation --- a double-blind initial submission of a full-length technical paper, a two-page SIGGRAPH Emerging Technologies abstract, a ``camera-ready'' journal article, a SIGCHI Extended Abstract, and more --- all by selecting the appropriate {\it template style} and {\it template parameters}.
+
+This document will explain the major features of the document class.
+For further information, the {\it \LaTeX\ User's Guide} is available from \url{https://www.acm.org/publications/proceedings-template}.
+
+\subsection{Template Styles}
+
+The primary parameter given to the ``\verb|acmart|'' document class is the {\it template style} which corresponds to the kind of publication or SIG publishing the work.
+This parameter is enclosed in square brackets and is a part of the {\verb|documentclass|} command:
+\begin{verbatim}
+  \documentclass[STYLE]{acmart}
+\end{verbatim}
+
+Journals use one of three template styles. All but three ACM journals use the {\verb|acmsmall|} template style:
+\begin{itemize}
+\item {\verb|acmsmall|}: The default journal template style.
+\item {\verb|acmlarge|}: Used by JOCCH and TAP.
+\item {\verb|acmtog|}: Used by TOG.
+\end{itemize}
+
+The majority of conference proceedings documentation will use the {\verb|acmconf|} template style.
+\begin{itemize}
+\item {\verb|acmconf|}: The default proceedings template style.
+\item{\verb|sigchi|}: Used for SIGCHI conference articles.
+\item{\verb|sigchi-a|}: Used for SIGCHI ``Extended Abstract'' articles.
+\item{\verb|sigplan|}: Used for SIGPLAN conference articles.
+\end{itemize}
+
+\subsection{Template Parameters}
 
@@
   \caption{1907 Franklin Model D roadster. Photograph by Harris \& Ewing, Inc. [Public domain], via Wikimedia Commons. (\url{https://goo.gl/VLCRBB}).}
-  \Description{The 1907 Franklin Model D roadster}
+  \Description{The 1907 Franklin Model D roadster.}
 \end{figure}
@@
 
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{references}
 \end{document}

### 2019-03-30 paper.tex
template

--- 2019-03-30 demo.tex
@@
 \ccsdesc[500]{Computing methodologies~Transfer learning}
+
+%
+% Keywords. The author(s) should pick words that accurately describe the work being
+% presented. Separate the keywords with commas.
 
@@
 
-% 1 Introdução
-% 1.1 Contribuições
-% 1.2 Visão Geral e Organização do Artigo
-% 1.3 Trabalhos Relacionados
-% 2 Método de Revisão Sistemática com Enfoque Meta Analítico
-% 2.1 o que é Enfoque Meta Analítico e por que adotá-lo
-% 2.2 
-% 3 Revisão da Literatura
-% 3.1 O que é Transferência de Aprendizado e por que é estudada
-% 3.1.1 Uma breve história da pesquisa na área
-% 3.3 Os "Clássicos"
-% 3.4 A "Fronteira"
-% 4 Problemas Abertos e Implicações para Projetos de Pesquisa
-% 5 Conclusão
-% 6 Referências
-
 \section{Introdução} 
@@
 \section{Conclusão}
-
-In addition to specifying the {\it template style} to be used in formatting your work, there are a number of {\it template parameters} which modify some part of the applied template style.
-A complete list of these parameters can be found in the {\it \LaTeX\ User's Guide.}
-
-Frequently-used parameters, or combinations of parameters, include:
-\begin{itemize}
-\item {\verb|anonymous,review|}: Suitable for a ``double-blind'' conference submission.
-Anonymizes the work and includes line numbers.
-Use with the \verb|\acmSubmissionID| command to print the submission's unique ID on each page of the work.
-\item{\verb|authorversion|}: Produces a version of the work suitable for posting by the author.
-\item{\verb|screen|}: Produces colored hyperlinks.
-\end{itemize}
-
-This document uses the following string as the first command in the source file: \verb|\documentclass[sigconf,screen]{acmart}|.
-
-\section{Modifications}
-
-Modifying the template --- including but not limited to: adjusting margins, typeface sizes, line spacing, paragraph and list definitions, and the use of the \verb|\vspace| command to manually adjust the vertical spacing between elements of your work --- is not allowed.
-
-{\bf Your document will be returned to you for revision if modifications are discovered.}
-
-\section{Typefaces}
-
-The ``\verb|acmart|'' document class requires the use of the ``Libertine'' typeface family.
-Your \TeX\ installation should include this set of packages.
-Please do not substitute other typefaces.
-The ``\verb|lmodern|'' and ``\verb|ltimes|'' packages should not be used, as they will override the built-in typeface families.
-
-\section{Title Information}
-
-The title of your work should use capital letters appropriately - \url{https://capitalizemytitle.com/} has useful rules for capitalization.
-Use the {\verb|title|} command to define the title of your work.
-If your work has a subtitle, define it with the {\verb|subtitle|} command.
-Do not insert line breaks in your title.
-
-If your title is lengthy, you must define a short version to be used in the page headers, to prevent overlapping text. The \verb|title| command has a ``short title'' parameter:
-\begin{verbatim}
-  \title[short title]{full title}
-\end{verbatim}
-
-\section{Authors and Affiliations}
-
-Each author must be defined separately for accurate metadata identification.
-Multiple authors may share one affiliation.
-Authors' names should not be abbreviated; use full first names wherever possible.
-Include authors' e-mail addresses whenever possible.
-
-Grouping authors' names or e-mail addresses, or providing an ``e-mail alias,'' as shown below, is not acceptable:
-\begin{verbatim}
-  \author{Brooke Aster, David Mehldau}
-  \email{dave,judy,steve@university.edu}
-  \email{firstname.lastname@phillips.org}
-\end{verbatim}
-
-The \verb|authornote| and \verb|authornotemark| commands allow a note to apply to multiple authors --- for example, if the first two authors of an article contributed equally to the work.
-
-If your author list is lengthy, you must define a shortened version of the list of authors to be used in the page headers, to prevent overlapping text.
-The following command should be placed just after the last \verb|\author{}| definition:
-\begin{verbatim}
-  \renewcommand{\shortauthors}{McCartney, et al.}
-\end{verbatim}
-Omitting this command will force the use of a concatenated list of all of the authors' names, which may result in overlapping text in the page headers.
-
-The article template's documentation, available at  \url{https://www.acm.org/publications/proceedings-template}, has a complete explanation of these commands and tips for their effective use.
-
-\section{Rights Information}
-
-Authors of any work published by ACM will need to complete a rights form.
-Depending on the kind of work, and the rights management choice made by the author, this may be copyright transfer, permission, license, or an OA (open access) agreement.
-
-Regardless of the rights management choice, the author will receive a copy of the completed rights form once it has been submitted.
-This form contains \LaTeX\ commands that must be copied into the source document.
-When the document source is compiled, these commands and their parameters add formatted text to several areas of the final document:
-\begin{itemize}
-\item the ``ACM Reference Format'' text on the first page.
-\item the ``rights management'' text on the first page.
-\item the conference information in the page header(s).
-\end{itemize}
-
-Rights information is unique to the work; if you are preparing several works for an event, make sure to use the correct set of commands with each of the works.
-
-\section{CCS Concepts and User-Defined Keywords}
-
-Two elements of the ``acmart'' document class provide powerful taxonomic tools for you to help readers find your work in an online search.
-
-The ACM Computing Classification System --- \url{https://www.acm.org/publications/class-2012} --- is a set of classifiers and concepts that describe the computing discipline.
-Authors can select entries from this classification system, via \url{https://dl.acm.org/ccs/ccs.cfm}, and generate the commands to be included in the \LaTeX\ source.
-
-User-defined keywords are a comma-separated list of words and phrases of the authors' choosing, providing a more flexible way of describing the research being presented.
-
-CCS concepts and user-defined keywords are required for all short- and full-length articles, and optional for two-page abstracts.
-
-\section{Sectioning Commands}
-
-Your work should use standard \LaTeX\ sectioning commands: \verb|section|, \verb|subsection|, \verb|subsubsection|, and \verb|paragraph|.
-They should be numbered; do not remove the numbering from the commands.
-
-Simulating a sectioning command by setting the first word or words of a paragraph in boldface or italicized text is {\bf not allowed.}
-
-\section{Tables}
-
-The ``\verb|acmart|'' document class includes the ``\verb|booktabs|'' package --- \url{https://ctan.org/pkg/booktabs} --- for preparing high-quality tables.
-
-Table captions are placed {\it above} the table.
-
-Because tables cannot be split across pages, the best placement for them is typically the top of the page nearest their initial cite.
-To ensure this proper ``floating'' placement of tables, use the environment \textbf{table} to enclose the table's contents and the table caption.
-The contents of the table itself must go in the \textbf{tabular} environment, to be aligned properly in rows and columns, with the desired horizontal and vertical rules.
-Again, detailed instructions on \textbf{tabular} material are found in the \textit{\LaTeX\ User's Guide}.
-
-Immediately following this sentence is the point at which Table~\ref{tab:freq} is included in the input file; compare the placement of the table here with the table in the printed output of this document.
-
-\begin{table}
-  \caption{Frequency of Special Characters}
-  \label{tab:freq}
-  \begin{tabular}{ccl}
-    \toprule
-    Non-English or Math&Frequency&Comments\\
-    \midrule
-    \O & 1 in 1,000& For Swedish names\\
-    $\pi$ & 1 in 5& Common in math\\
-    \$ & 4 in 5 & Used in business\\
-    $\Psi^2_1$ & 1 in 40,000& Unexplained usage\\
-  \bottomrule
-\end{tabular}
-\end{table}
-
-To set a wider table, which takes up the whole width of the page's live area, use the environment \textbf{table*} to enclose the table's contents and the table caption.
-As with a single-column table, this wide table will ``float'' to a location deemed more desirable.
-Immediately following this sentence is the point at which Table~\ref{tab:commands} is included in the input file; again, it is instructive to compare the placement of the table here with the table in the printed output of this document.
-
-\begin{table*}
-  \caption{Some Typical Commands}
-  \label{tab:commands}
-  \begin{tabular}{ccl}
-    \toprule
-    Command &A Number & Comments\\
-    \midrule
-    \texttt{{\char'134}author} & 100& Author \\
-    \texttt{{\char'134}table}& 300 & For tables\\
-    \texttt{{\char'134}table*}& 400& For wider tables\\
-    \bottomrule
-  \end{tabular}
-\end{table*}
-
-\section{Math Equations}
-You may want to display math equations in three distinct styles:
-inline, numbered or non-numbered display.  Each of
-the three are discussed in the next sections.
-
-\subsection{Inline (In-text) Equations}
-A formula that appears in the running text is called an
-inline or in-text formula.  It is produced by the
-\textbf{math} environment, which can be
-invoked with the usual \texttt{{\char'134}begin\,\ldots{\char'134}end}
-construction or with the short form \texttt{\$\,\ldots\$}. You
-can use any of the symbols and structures,
-from $\alpha$ to $\omega$, available in
-\LaTeX~\cite{Lamport:LaTeX}; this section will simply show a
-few examples of in-text equations in context. Notice how
-this equation:
-\begin{math}
-  \lim_{n\rightarrow \infty}x=0
-\end{math},
-set here in in-line math style, looks slightly different when
-set in display style.  (See next section).
-
-\subsection{Display Equations}
-A numbered display equation---one set off by vertical space from the
-text and centered horizontally---is produced by the \textbf{equation}
-environment. An unnumbered display equation is produced by the
-\textbf{displaymath} environment.
-
-Again, in either environment, you can use any of the symbols
-and structures available in \LaTeX\@; this section will just
-give a couple of examples of display equations in context.
-First, consider the equation, shown as an inline equation above:
-\begin{equation}
-  \lim_{n\rightarrow \infty}x=0
-\end{equation}
-Notice how it is formatted somewhat differently in
-the \textbf{displaymath}
-environment.  Now, we'll enter an unnumbered equation:
-\begin{displaymath}
-  \sum_{i=0}^{\infty} x + 1
-\end{displaymath}
-and follow it with another numbered equation:
-\begin{equation}
-  \sum_{i=0}^{\infty}x_i=\int_{0}^{\pi+2} f
-\end{equation}
-just to demonstrate \LaTeX's able handling of numbering.
-
-\section{Figures}
-
-The ``\verb|figure|'' environment should be used for figures.
-One or more images can be placed within a figure.
-If your figure contains third-party material, you must clearly identify it as such, as shown in the example below.
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{sample-franklin}
-  \caption{1907 Franklin Model D roadster. Photograph by Harris \& Ewing, Inc. [Public domain], via Wikimedia Commons. (\url{https://goo.gl/VLCRBB}).}
-  \Description{The 1907 Franklin Model D roadster}
-\end{figure}
-
-Your figures should contain a caption which describes the figure to the reader.
-Figure captions go below the figure.
-Your figures should {\bf also} include a description suitable for screen readers, to assist the visually-challenged to better understand your work.
-
-Figure captions are placed {\it below} the figure.
-
-\subsection{The ``Teaser Figure''}
-
-A ``teaser figure'' is an image, or set of images in one figure, that are placed after all author and affiliation information, and before the body of the article, spanning the page.
-If you wish to have such a figure in your article, place the command immediately before the \verb|\maketitle| command:
-\begin{verbatim}
-  \begin{teaserfigure}
-    \includegraphics[width=\textwidth]{sampleteaser}
-    \caption{figure caption}
-    \Description{figure description}
-  \end{teaserfigure}
-\end{verbatim}
 
@@
 
-\section{Acknowledgments}
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{reference}
 
-Identification of funding sources and other support, and thanks to individuals and groups that assisted in the research and the preparation of the work should be included in an acknowledgment section, which is placed just before the reference section in your document.
-
-This section has a special environment:
-\begin{verbatim}
-  \begin{acks}
-  ...
-  \end{acks}
-\end{verbatim}
-so that the information contained therein can be more easily collected during the article metadata extraction phase, and to ensure consistency in the spelling of the section heading.
-
-Authors should not prepare this section as a numbered or unnumbered {\verb|\section|}; please use the ``{\verb|acks|}'' environment.
-
-\section{Appendices}
-
-If your work needs an appendix, add it before the ``\verb|\end{document}|'' command at the conclusion of your source document.
-
-Start the appendix with the ``\verb|appendix|'' command:
-\begin{verbatim}
-  \appendix
-\end{verbatim}
-and note that in the appendix, sections are lettered, not numbered. This document has two appendices, demonstrating the section and subsection identification method.
-
-\section{SIGCHI Extended Abstracts}
-
-The ``\verb|sigchi-a|'' template style (available only in \LaTeX\ and not in Word) produces a landscape-orientation formatted article, with a wide left margin.
-Three environments are available for use with the ``\verb|sigchi-a|'' template style, and produce formatted output in the margin:
-\begin{itemize}
-\item {\verb|sidebar|}:  Place formatted text in the margin.
-\item {\verb|marginfigure|}: Place a figure in the margin.
-\item {\verb|margintable|}: Place a table in the margin.
-\end{itemize}
-
-%
-% The acknowledgments section is defined using the "acks" environment (and NOT an unnumbered section). This ensures
-% the proper identification of the section in the article metadata, and the consistent spelling of the heading.
-\begin{acks}
-To Robert, for the bagels and explaining CMYK and color spaces.
-\end{acks}
-
-%
-% The next two lines define the bibliography style to be used, and the bibliography file.
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{sample-base}
-
-% 
-% If your work has an appendix, this is the place to put it.
-\appendix
-
-\section{Research Methods}
-
-\subsection{Part One}
-
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Morbi malesuada, quam in pulvinar varius, metus nunc fermentum urna, id sollicitudin purus odio sit amet enim.
-Aliquam ullamcorper eu ipsum vel mollis.
-Curabitur quis dictum nisl.
-Phasellus vel semper risus, et lacinia dolor.
-Integer ultricies commodo sem nec semper.
-
-\subsection{Part Two}
-
-Etiam commodo feugiat nisl pulvinar pellentesque.
-Etiam auctor sodales ligula, non varius nibh pulvinar semper.
-Suspendisse nec lectus non ipsum convallis congue hendrerit vitae sapien.
-Donec at laoreet eros.
-Vivamus non purus placerat, scelerisque diam eu, cursus ante.
-Etiam aliquam tortor auctor efficitur mattis.
-
-\section{Online Resources}
-
-Nam id fermentum dui.
-Suspendisse sagittis tortor a nulla mollis, in pulvinar ex pretium.
-Sed interdum orci quis metus euismod, et sagittis enim maximus.
-Vestibulum gravida massa ut felis suscipit congue.
-Quisque mattis elit a risus ultrices commodo venenatis eget dui.
-Etiam sagittis eleifend elementum.
-
-Nam interdum magna at lectus dignissim, ac dignissim lorem rhoncus. Maecenas eu arcu ac neque placerat aliquam. Nunc pulvinar massa et mattis lacinia.
-
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{references}
 \end{document}

### 2019-03-30 paper.tex
abstract feito

@@
 % The "title" command has an optional parameter, allowing the author to define a "short title" to be used in page headers.
-\title[Fronteiras da Transferência de Aprendizagem: uma revisão sistemática com enfoque meta-analítico]{Fronteiras da Transferência de Aprendizagem: \\
+\title[Fronteiras da Transferência de Aprendizado: uma revisão sistemática com enfoque meta-analítico]{Fronteiras da Transferência de Aprendizado: \\
 uma revisão sistemática com enfoque meta-analítico}
@@
 \begin{abstract}
-A clear and well-documented \LaTeX\ document is presented as an article formatted for publication by ACM in 
-a conference proceedings or journal publication. Based on the ``acmart'' document class, this article presents
-and explains many of the common variations, as well as many of the formatting elements
-an author may use in the preparation of the documentation of their work.
+  Humanos e animais conseguem aprender com poucas amostras \cite{goodfellow} e apresentam extraordinária capacidade de generalização que os algoritmos de aprendizagem de máquina ainda estão longe de alcançar.
+Os modelos mais bem sucedidos da atualidade exigem uma enormidade de dados bem rotulados que são caros e difíceis de obter, tornando-se hoje um dos maiores empecilhos para aplicações práticas.
+Esse cenário prova o grande potencial da área de Transferência de Aprendizado, que tem por objetivo aproveitar o conhecimento obtido em uma atividade para aprender mais eficientemente outras, que guardem alguma relação com a primeira.
+O presente estudo visa apresentar uma revisão sistemática da literatura e identificar, com embasamento quantitativo, as principais contribuições para a área.
+Além disso, usamos o acoplamento bibliográfico para identificar trabalhos na fronteira do conhecimento e fizemos uma análise textual, em resumos e palavras-chave, comparando estes com os "clássicos" da área de forma a mapear para que direção a pesquisa avança.
+
 \end{abstract}
+
+%   O objetivo deste estudo é mostrar uma revisão sistemática da literatura com as principais contribuições em transferência de aprendizagem.
+Uma pesquisa exploratória foi realizada, de abordagem quantitativa, por meio da Teoria do Enfoque Meta Analítico Consolidado
+
+% Esta pesquisa quer responder: quais são as fronteiras do conhecimento em transferência de aprendizagem?  É possível embasar essa avaliação em dados bibliométricos?
+
+%   Os melhores casos de sucesso foram desenvolvidos exclusiva
 
@@
   \subsection{Sumarização}
-\subsection{Revisão da Literatura}
+\section{Revisão da Literatura}
   \subsection{Transferência de Aprendizagem}
@@
 \section{Conclusão}
-
-\section{Citations and Bibliographies}
-
-The use of \BibTeX\ for the preparation and formatting of one's references is strongly recommended.
-Authors' names should be complete --- use full first names (``Donald E.
-Knuth'') not initials (``D.
-E.
-Knuth'') --- and the salient identifying features of a reference should be included: title, year, volume, number, pages, article DOI, etc.
-
-The bibliography is included in your source document with these two commands, placed just before the \verb|\end{document}| command:
-\begin{verbatim}
 \bibliographystyle{ACM-Reference-Format}
-  \bibliography{bibfile}
-\end{verbatim}
-where ``\verb|bibfile|'' is the name, without the ``\verb|.bib|'' suffix, of the \BibTeX\ file.
-
-Citations and references are numbered by default.
-A small number of ACM publications have citations and references formatted in the ``author year'' style; for these exceptions, please include this command in the {\bf preamble} (before ``\verb|\begin{document}|'') of your \LaTeX\ source:
-\begin{verbatim}
-  \citestyle{acmauthoryear}
-\end{verbatim}
-
-Some examples.
-A paginated journal article \cite{Abril07}, an enumerated journal article \cite{Cohen07}, a reference to an entire issue \cite{JCohen96}, a monograph (whole book) \cite{Kosiur01}, a monograph/whole book in a series (see 2a in spec.
-document)
-\cite{Harel79}, a divisible-book such as an anthology or compilation \cite{Editor00} followed by the same example, however we only output the series if the volume number is given \cite{Editor00a} (so Editor00a's series should NOT be present since it has no vol.
-no.),
-a chapter in a divisible book \cite{Spector90}, a chapter in a divisible book in a series \cite{Douglass98}, a multi-volume work as book \cite{Knuth97}, an article in a proceedings (of a conference, symposium, workshop for example) (paginated proceedings article) \cite{Andler79}, a proceedings article with all possible elements \cite{Smith10}, an example of an enumerated proceedings article \cite{VanGundy07}, an informally published work \cite{Harel78}, a doctoral dissertation \cite{Clarkson85}, a master's thesis: \cite{anisi03}, an online document / world wide web resource \cite{Thornburg01, Ablamowicz07, Poker06}, a video game (Case 1) \cite{Obama08} and (Case 2) \cite{Novak03} and \cite{Lee05} and (Case 3) a patent \cite{JoeScientist001}, work accepted for publication \cite{rous08}, 'YYYYb'-test for prolific author \cite{SaeediMEJ10} and \cite{SaeediJETC10}.
-Other cites might contain 'duplicate' DOI and URLs (some SIAM articles) \cite{Kirschmer:2010:AEI:1958016.1958018}.
-Boris / Barbara Beeton: multi-volume works as books \cite{MR781536} and \cite{MR781537}.
-A couple of citations with DOIs: \cite{2004:ITE:1009386.1010128,Kirschmer:2010:AEI:1958016.1958018}.
-Online citations: \cite{TUGInstmem, Thornburg01, CTANacmart}.
-
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{reference}
-
+\bibliography{references}
 \end{document}

### 2019-04-03 paper.tex
ajustando references
versão entrega Li

@@
 \documentclass[sigconf]{acmart}
-\usepackage{graphicx}
+\usepackage{graphicx, wrapfig, floatflt}
 \graphicspath{{imgs/}}
+\usepackage[para]{footmisc} %  footnotes na mesma linha
+\usepackage{stfloats}
+\usepackage{enumerate}
+
+\usepackage[most]{tcolorbox}
 \usepackage[english,ngerman,brazilian]{babel}
+
+\newtcolorbox[blend into=figures]{card}[2][]{enhanced,
+  float=tbp,title={#2},
+  colframe=gray!75!black,olback=yellow!5!white,#1
+}
+
 \settopmatter{printacmref=false}
@@
 \pagestyle{plain}
-
+\captionsetup{justification   = raggedright,
+              singlelinecheck = false}
+\newcommand{\source}[2]{\raggedleft{}\vspace*{-7mm}\caption*{ \textmd{\scriptsize{Dados: {#1}.\hfill Ferramenta:{#2}}}}}
 \def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08emT\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
@@
 %
-% The "author" command and its associated commands are used to define the authors and their affiliations.
-% Of note is the shared affiliation of the first two authors, and the "authornote" and "authornotemark" commands
-% used to denote shared contribution to the research.
+
 \author{Fred Guth}
-% \authornote{Both authors contributed equally to this research.}
 \email{fredguth@fredguth.com}
@@
 
-%
-% By default, the full list of authors will be used in the page headers. Often, this list is too long, and will overlap
-% other information printed in the page headers. This command allows the author to define a more concise list
-% of authors' names for this purpose.
-\renewcommand{\shortauthors}{Guth,.}
+\renewcommand{\shortauthors}{Guth, F.}
 
-%
-% The abstract is a short summary of the work to be presented in the article.
 \begin{abstract}
-  Humanos e animais conseguem aprender com poucas amostras \cite{goodfellow} e apresentam extraordinária capacidade de generalização que os algoritmos de aprendizagem de máquina ainda estão longe de alcançar.
+  Humanos e animais conseguem aprender com poucas amostras e apresentam extraordinária capacidade de generalização que os algoritmos de aprendizagem de máquina ainda estão longe de alcançar~\cite{goodfellow}.
 Os modelos mais bem sucedidos da atualidade exigem uma enormidade de dados bem rotulados que são caros e difíceis de obter, tornando-se hoje um dos maiores empecilhos para aplicações práticas.
-Esse cenário prova o grande potencial da área de Transferência de Aprendizado, que tem por objetivo aproveitar o conhecimento obtido em uma atividade para aprender mais eficientemente outras, que guardem alguma relação com a primeira.
+Tais fatos apontam para o grande potencial da área de Transferência de Aprendizado, que tem por objetivo aproveitar o conhecimento obtido em uma atividade para aprender mais eficientemente outras, que guardem alguma relação com a primeira.
 O presente estudo visa apresentar uma revisão sistemática da literatura e identificar, com embasamento quantitativo, as principais contribuições para a área.
-Além disso, usamos o acoplamento bibliográfico para identificar trabalhos na fronteira do conhecimento e fizemos uma análise textual, em resumos e palavras-chave, comparando estes com os "clássicos" da área de forma a mapear para que direção a pesquisa avança.
-
+Além disso, usamos medidas de acoplamento bibliográfico para identificar trabalhos na fronteira do conhecimento e fizemos uma análise textual, em resumos e palavras-chave, comparando estes com os "clássicos" ~da área de forma a mapear para que direção a pesquisa avança.
 \end{abstract}
-
-%   O objetivo deste estudo é mostrar uma revisão sistemática da literatura com as principais contribuições em transferência de aprendizagem.
-Uma pesquisa exploratória foi realizada, de abordagem quantitativa, por meio da Teoria do Enfoque Meta Analítico Consolidado
-
-% Esta pesquisa quer responder: quais são as fronteiras do conhecimento em transferência de aprendizagem?  É possível embasar essa avaliação em dados bibliométricos?
-
-%   Os melhores casos de sucesso foram desenvolvidos exclusiva
-
-%
-% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
-% Please copy and paste the code instead of the example below.
-%
 
@@
 \end{CCSXML}
-
 \ccsdesc[500]{Computing methodologies~Transfer learning}
-
-%
-% Keywords. The author(s) should pick words that accurately describe the work being
-% presented. Separate the keywords with commas.
 
@@
 
-%
-% A "teaser" image appears between the author and affiliation information and the body 
-% of the document, and typically spans the page. 
-% \begin{teaserfigure}
-%   \includegraphics[width=\textwidth]{sampleteaser}
-%   \caption{Seattle Mariners at Spring Training, 2010.}
-%   \Description{Enjoying the baseball game from the third-base seats. Ichiro Suzuki preparing to bat.}
-%   \label{fig:teaser}
-% \end{teaserfigure}
-
-%
-% This command processes the author and affiliation and title information and builds
-% the first part of the formatted document.
 \maketitle
@@
 \section{Introdução} 
+  Recentes avanços em aprendizagem de máquina tornam possíveis aplicações que são capazes de reconhecer pessoas, lugares e objetos com acurácia super-humana\cite{fei}, diagnosticar câncer de pele tão bem quanto dermatologistas\cite{skin_cancer},  ver através de paredes usando sinais de rádio\cite{wifi}, entre tantas outras.
+Apesar de tanto sucesso, os modelos mais bem sucedidos da atualidade exigem uma enormidade de dados bem rotulados que são caros e difíceis de obter, pois, em geral, o jeito padrão de se treinar modelos é sempre começar \emph{tabula rasa}, ou seja, com uma inicialização aleatória dos parâmetros~\cite{Ruder2019Neural}.
+
+  Aprender assim, a partir do nada, é contrário à forma como os humanos o fazem.
+Em nosso dia a dia, transferimos conhecimento a todo momento.
+Saber tocar piano facilita aprender tocar órgão.
+Saber português ajuda a aprender espanhol.
+Pessoas conseguem inteligentemente aplicar conhecimento prévio para resolver novos problemas com maior eficácia e eficiência\cite{PanYang}.
+Algoritmos de aprendizagem de máquina ainda estão longe de alcançar essa extraordinária capacidade de generalização\cite{goodfellow}.
+Estudos recentes~\cite{DBLP:journals/corr/JiaL17} mostram que os algoritmos atuais não generalizam bem além de dados vistos durante o treinamento.
+  \begin{figure}[h]
+    \includegraphics[width=\columnwidth]{citacoes_por_ano}
+    \source{Web of Science (março/2019)}{Excel}
+    \caption{Evolução do número de citações aos artigos em transferência de aprendizado nos últimos 10 anos. Até 2020, o número de citações aos artigos na área deverá dobrar.}
+    \label{fig:citacoes_por_ano}
+  \end{figure}
+
+  Tais fatos apontam  o grande potencial ainda inalcançado da área de Transferência de Aprendizado (TL\footnote{Do inglês, \emph{Transfer Learning}.}), que tem por objetivo aproveitar o conhecimento obtido em uma atividade para aprender mais eficientemente outras, que guardem alguma relação com a primeira.
+
+  Na prática, entretanto, TL ainda é tratada de uma forma \textit{ad hoc}, sendo os métodos de transferências meras extensões dos algoritmos de aprendizado utilizados \cite{torrey}.
+Essa dicotomia entre a importância do problema e a inexistência de práticas e teorias consolidadas, tornam TL um campo promissor e interessante para pesquisa.
+
+  \begin{quote} "Transferência de aprendizado será o próximo motor do sucesso comercial com aprendizado de máquinas." \hfill ---Andrew Ng, Tutorial NIPS 2016 \cite{ANg}
+  \end{quote}
+
+  Portanto, não é de se estranhar o crescente interesse pelo assunto (vide figuras \ref{fig:citacoes_por_ano} e \ref{fig:artigos_por_ano}).
+
+  \subsection{Objetivo}\label{objetivos}
+   Esta pesquisa quer responder duas perguntas:
+    \begin{enumerate}
+      \item{Quais são as fronteiras do conhecimento em transferência de aprendizado?}
+      \item {É possível embasar essa avaliação em dados bibliométricos?}
+    \end{enumerate}
+    Para responder estas perguntas é preciso, antes, analisar a literatura do tema, revelar as principais contribuições na área e como se relacionam.
+
   \subsection{Contribuições}
+
+    \begin{enumerate}[C1.]
+      \item Apresentamos uma revisão sistemática atualizada da literatura em Transferência de Aprendizagem, usando a abordagem TEMAC (\S \ref{TEMAC}).
+Esse método nos permite focar nas contribuições de alto impacto.
+      \item Estendemos o TEMAC com uma análise textual de resumos (\S \ref{detalha}\ref{analiseTextual}), utilizando o \emph{ScatterText}~\cite{kessler2017scattertext}, o que até onde sabemos, é uma aplicação original desta ferramenta de visualização.
+      \item Apontamos com embasamento quantitativo as direções das frentes de pesquisa da área e apontamos lacunas e problemas em aberto.
+    \end{enumerate}
+
   \subsection{Visão Geral e Organização do Artigo}
-  \subsection{Trabalhos Relacionados}
-\section{Método de Revisão Sistemática}
-  \subsection{O Enfoque Meta Analítico}
-  \subsection{Análise de Co-citações}
-  \subsection{Análise de Acoplamento Bibliográfico}
-  \subsection{Análise Textual}
-  \subsection{Sumarização}
-\section{Revisão da Literatura}
-  \subsection{Transferência de Aprendizagem}
+  Ainda nesta breve introdução apresentaremos os \emph{Trabalhos Relacionados} (\S\ref{related}).
+Na próxima seção (\S\ref{TEMAC}), apresentaremos o método de pesquisa que dá embasamento quantitativo que queremos para atingirmos nossos objetivos (\S\ref{objetivos}).
+Na seção \S\ref{literatura}, apresentamos os resultados, que são a própria  \emph{Revisão da Literatura}.
+\emph{Problemas em Aberto} são discutidos na seção \S\ref{emAberto}.
+Por fim, concluímos o trabalho (\S\ref{conclusao}) apresentando as respostas às perguntas de \S\ref{objetivos}.
+
+  \subsection{Trabalhos Relacionados}\label{related}
+  Pan, S.
+(2009)\cite{PanYang} é a mais citada revisão sobre \emph{Transferência de Aprendizado}.
+Hohman, F.
+(2018)\cite{hohman2018visual} faz uma revisão sistemática da literatura sobre análises visuais em \emph{deep learning} e utiliza um método bastante interessante baseado nos cinco W's e um H (Why, Who, What, How, When, and Where).
+Ruder, S.
+(2019) \cite{Ruder2019Neural} apresenta um capítulo com uma revisão bibliográfica e taxonomia para o contexto de NLP.
+Park, I.
+(2018)\cite{park2018identifying} utiliza uma abordagem meta-analítica para identificar a fronteira de pesquisa em \emph{Pattern Recognition}.
+\section{Método: Revisão com Enfoque Meta-Analítico Consolidado}\label{TEMAC}
+Na presente pequisa utilizamos o método de revisão sistemática da Teoria do Enfoque Meta-Analítico (TEMAC) ~\cite{Mariano}, que visa oferecer um embasamento quantitativo para a escolha da literatura.
+Apesar do nome, TEMAC não deve ser confundida com \textbf{meta-análise}, pois o foco desta é gerar conhecimento por meio de dados empíricos secundários, enquanto o daquela é oferecer uma sistematização da escolha bibliográfica.
+
+A abordagem TEMAC é dividida em 3 etapas: 
+\begin{enumerate}[a)]
+  \item preparação da pesquisa;
+  \item apresentação e inter-relação dos dados;
+  \item detalhamento, modelo integrador e validação por evidências.
+\end{enumerate}
+
+\subsection{Preparação da Pesquisa}
+Foi realizada uma busca na base de dados \emph{Clarivate Analyticis Web of Science (WoS)}, conforme especificado na figura \ref{card:wos}, no dia 31 de março de 2019, que resultou em 1.289 artigos encontrados.
+Nota-se que o tema é de crescente interesse (figura \ref{fig:artigos_por_ano}) e projeta-se que em 3 anos o número de artigos irá praticamente dobrar (crescimento exponencial com $R^2=0,9789$).
+\begin{figure}[htp]
+
+\begin{tcolorbox}[colback=yellow!5!white,colframe=gray!75!black,title={Results: 1,289 (from Web of Science Core Collection)}]
+  \footnotesize{
+    \begin{verbatim}
+    You searched for: 
+    TOPIC: ("transfer learning")
+    Refined by: 
+    WEB OF SCIENCE CATEGORIES: 
+      ( COMPUTER SCIENCE ARTIFICIAL INTELLIGENCE )
+    AND LANGUAGES: ( ENGLISH ) 
+    AND RESEARCH AREAS: ( COMPUTER SCIENCE )
+    Timespan: 2009-2019. 
+    Indexes: SCI-EXPANDED, CPCI-S, ESCI.
+    \end{verbatim}
+  }
+
+\end{tcolorbox}
+\caption{"Busca10anos": Parâmetros de busca na base \emph{Web of Science}.}
+\label{card:wos}
+\end{figure}
+
+Também fizemos uma busca mais restrita aos trabalhos mais recentes (figura \ref{card:sota}) para a análise das \emph{Fronteiras de Pesquisa} (\S\ref{fronteiras}).
+\begin{figure}[htp]
+  \begin{tcolorbox}[colback=yellow!5!white,colframe=gray!75!black,title={Results: 384 (from Web of Science Core Collection)}]
+    \footnotesize{
+  \begin{verbatim}
+  You searched for: 
+  TOPIC: ("transfer learning")
+  Refined by: 
+  WEB OF SCIENCE CATEGORIES: 
+    ( COMPUTER SCIENCE ARTIFICIAL INTELLIGENCE )
+  AND LANGUAGES: ( ENGLISH ) 
+  AND RESEARCH AREAS: ( COMPUTER SCIENCE )
+  AND PUBLICATION YEARS: ( 2019 OR 2018 OR 2017 )
+  AND DOCUMENT TYPES: ( PROCEEDINGS PAPER )
+  Timespan: 2009-2019. 
+  Indexes: SCI-EXPANDED, CPCI-S, ESCI.
+  \end{verbatim}
+    }
+
+  \end{tcolorbox}
+  \caption{"Busca3anos"Parâmetros de busca para análise de Fronteira.}
+  \label{card:sota}
+\end{figure}
+
+\subsection{Apresentação e Inter-relação dos Dados}\label{inter}
+Nessa etapa analisamos (vide \S\ref{sec:panorama}):
+\begin{enumerate}
+  \item artigos mais citados e com maior média de citações por ano;
+  \item evolução ano a ano do número de artigos;
+  \item evolução ano a ano do número de citações;
+  \item autores com mais artigos e com mais citações;
+  \item conferência com mais artigos e citações;
+  \item instituições com mais artigos e citações;
+  \item países que mais produzem artigos científicos na área;
+  \item frequência de palavras-chave (figuras \ref{fig:clouds} e \ref{fig:scatterText}).
+\end{enumerate}
+\subsection{Detalhamento e Validação}\label{detalha}
+
+\begin{enumerate}[a)]
+  \item{\textbf{Análise de co-citações}: Co-citação é definida como a frequência em que dois documentos são citados em uma mesma lista de referências.
+Assume-se que são "peças"~que compõem uma mesma "estrutura de conhecimento".
+A análise de co-citações, portanto, é útil para mapear a herança intelectual de um determinado campo de estudos sob a ótica das publicações de alto-impacto, mas tende a negligenciar a dinâmica na fronteira de pesquisa\cite{Vogel2012}, justamente porque os trabalhos na fronteira não tiveram tempo para serem citados e pontuar nas métricas de impacto.\\Para a análise de co-citações, utilizamos o software livre \emph{VosViewer 1.6.10}~\cite{VOSviewer}, que \emph{clusterizou} citações das listas de referências dos artigos da "Busca10anos" (o que abrange artigos que não fazem parte do resultado da busca).
+Foram identificados 3 núcleos de conhecimento (figura \ref{fig:classicos}).
+  }
+
+  \item{\textbf{Análise de acoplamento bibliográfico}: Acoplamento bibliográfico ocorre quando dois documentos tem pelo menos um documento em comum, ou seja, documentos acoplam se suas bibliografias se sobrepõem.
+Como existe uma ordem cronológica entre o documento que cita e o que é citado, o acoplamento bibliográfico permite traçar gerações de pesquisas e com isso identificar as que estão na fronteira do conhecimento.
+É importante ressaltar que, neste contexto, estar na fronteira não é necessariamente ser promissor.
+Essa é uma limitação da abordagem quantitativa, por isso é preciso complementá-la com uma análise qualitativa para propor, ainda que com certo grau de subjetividade, os trabalhos promissores a se tornarem os "clássicos de amanhã".
+\\Pelo TEMAC, a análise de acoplamento bibliográfico deve ser feita para o período um período não superior aos últimos 3 anos.
+Em nossa análise, escolhemos o período entre 2017 e março de 2019 e nos restringimos também aos artigos de conferência  (figura \ref{card:sota}), assumindo que publicação em revista têm uma maior tempo de revisão, portanto, artigos em conferências são publicados mais proximamente ao momento de sua confecção e, portanto, mais à frente.
+  }
+  \item{\textbf{Análise textual (\emph{bag-of-words}  com tf-idf)}: nesta análise, tratamos os artigos como \emph{bag-of-words} e usamos o conceito de \textbf{tf-idf} para definir as palavras (termos) que melhor identificam determinados artigos (documentos).
+Por exemplo, quais palavras melhor explicam trabalhos de fronteira \emph{vis-à-vis} os que explicam os documentos resultantes da "Busca10anos" (vide figura \ref{fig:scatterText}).
+A métrica \textbf{tf-idf} pode ser definida como:
+  \begin{equation}
+    \mathrm{tfidf}(t,d,D) = \mathrm{tf}(t,d) \cdot \mathrm{idf}(t, D)
+  \end{equation}
+  onde $\mathrm{tf}(t,d)$ é a frequência em que um termo $t$ aparece em um documento $d$ e $\mathrm{idf}(t, D)$ é o inverso da frequência do termo $t$ no conjunto de documentos $D$ (o \emph{corpus}). 
+  \begin{equation}
+    \mathrm{tf}(t,d) = 0.5 + 0.5 \cdot  \frac{f_{t, d}}{\max\{f_{t', d}:t' \in d\}}
+  \end{equation}
+  \begin{equation}
+    \mathrm{idf}(t, D) =  \log \frac{N}{|\{d \in D: t \in d\}|}
+  \end{equation}
+  onde:
+  \begin{description}
+    \item $N$: número de documentos do corpus $N = {|D|}$
+   \item$ |\{d \in D: t \in d\}| $ : número de documentos onde o termo  $ t $ aparece (i.e., $ \mathrm{tf}(t,d) \neq 0$).
+Para evitar uma divisão por zero caso o termo não esteja no corpus, é como fazer um ajuste no denominador para: $1 + |\{d \in D: t \in d\}|$.
+  \end{description}
+
+  A métrica \textbf{tf-idf} é a base da ferramenta visual \emph{ScatterText}~\cite{kessler2017scattertext} que foi utilizada para gerar as figuras \ref{fig:scatterText_documentbased} e \ref{fig:scatterText}.
+
+  }\label{analiseTextual}
+\end{enumerate}
+
+\section{Revisão da Literatura}\label{literatura}
   \subsection{Um breve histórico}
-  \subsection{Os Clássicos}
-  \subsection{A Fronteira}
-\section{Problemas em Aberto}
-\section{Conclusão}
+  Pesquisa em transferência de aprendizado tem atraído mais e mais atenção desde 1995, quando foi tema de  um workshop na NIPS-95 que discutiu a necessidade de métodos de aprendizado de máquina que retém e reusam conhecimento previamente obtido~\cite{PanYang}.
+
+  Desde então, ainda que com diferentes nomes (\emph{learning to learn, life-long learning, knowledge transfer} entre outros), a área vem recebendo cada vez mais atenção (figuras \ref{fig:citacoes_por_ano} e \ref{fig:artigos_por_ano}).
+Em 2005 um anúncio de um projeto da DARPA foi talvez a primeira menção que ao termo \emph{transfer learning}, definindo-o como o objetivo de extrair conhecimento de um ou mais \emph{tarefas fonte} e aplicá-lo a \emph{tarefas alvo}~\cite{PanYang}.
+Se removermos a restrição de data de início da "Busca10Anos"(figura \ref{card:wos}), confirmamos que os primeiros artigos com o termo "transfer learning" são de 2005.
+
+  \begin{figure}[h]
+    \includegraphics[width=\columnwidth]{artigos_por_ano}
+    \source{Web of Science (março/2019)}{Excel}
+    \caption{Evolução do número de artigos em Tranferência de Aprendizado nos últimos 10 anos. Até 2021, o número de artigos na área deverá dobrar.}
+    \label{fig:artigos_por_ano}
+  \end{figure}
+
+  Em 2012, a rede neural profunda\footnote{deep neural network} submetida pelo time de Alex Krizhevsky et.
+al.
+para o desafio Imagenet (ILSVRC) performou 41\% melhor que o segundo colocado, dando início ao crescimento explosivo de \emph{deep learning}.
+Tal sucesso ressaltou a importância dos dados para o avanço de inteligência artificial, mas também permitiu uma igual quebra de paradigma em transferência de aprendizado: percebeu-se que modelos treinados na ImageNet podiam facilmente ser usados para inicializar outros modelos e reduzir em muito a necessidade de amostras~\cite{Ruder2019Neural}.
+Essa abordagem de "sintonia fina"\footnote{\emph{fine-tuning}} permitiu resultados bons em diversas tarefas com ordens de magnitude a menos de dados (vide \S\ref{terceira_onda}).
+
+  Atualmente, \emph{transferência de aprendizado} é um tópico comum em prestigiadas conferências como CVPR, ICCV, ICPR e NeurIPS (vide figura \ref{fig:toptop}, \emph{Top 10 encontros}).
+  \subsection{Notações e definições}
+  \begin{figure}[h]
+    \fbox{\includegraphics[width=\columnwidth]{taxonomy.pdf}}
+    \caption{Visão geral dos diversos tipos de transferência. Taxonomia de \emph{A Survey on Transfer Learning}\protect{~\cite{PanYang}} }
+    \label{fig:taxonomia}
+  \end{figure}
+    Transferência de aprendizado envolve os conceitos de \emph{domínio} e \emph{tarefa}.
+Seguindo a notação de \cite{PanYang}, um domínio $\mathcal{D}$ é composto de um espaço de características $\mathcal{X}\subset R^d$ e uma distribuição marginal $P(\mathrm{X})$, onde $\mathrm{S}=\{x_1, \dots, x_n\}\in\mathcal{X} $.
+Para um problema de classificação de imagens, por exemplo, $\mathcal{X}$ é o espaço de todas possíveis representações de imagens, com suas dimensões e canais, $x_i$ é uma imagem e $\mathrm{S}$ é o \textit{dataset} de treinamento.
+
+    Dado um domínio $\mathcal{D}=\{\mathcal{X}, P(X)\}$, uma tarefa $\mathcal{T}$ é definida pelo espaço de rótulos $\mathcal{Y}$ com distribuição condicional $P(\mathrm{Y}|\mathrm{X})$, ou seja, $\mathcal{T}=\{\mathcal{Y}, f(\cdot)\}$, onde $f(\cdot)$ é uma função objetivo que dado um $x_i \in \mathcal{D}$, prediz seu correspondente $y_i \in \mathcal{Y}$.
+
+    Dado um domínio fonte $\mathcal{D_S}$ e uma tarefa $\mathcal{T_S}$, um domínio destino $\mathcal{D_T}$ e uma tarefa $\mathcal{T_T}$, \textbf{transferência de aprendizado} objetiva auxiliar o aprendizado da função de predição $f_T(\cdot)$ em  $\mathcal{D_T}$ usando conhecimento de $\mathcal{D_S}$ e $\mathcal{T_S}$, onde $\mathcal{D_S}\neq\mathcal{D_T}$ ou $\mathcal{T_S}\neq\mathcal{T_T}$.
+
+    No presente artigo, usaremos a taxonomia apresentada por \cite{PanYang} (figura \ref{fig:taxonomia}). 
+    \subsection{O panorama da produção científica na área}\label{sec:panorama}
+  Na figura \ref{fig:toptop}, Top 10 autores, é possível notar que o autor com mais citações é Pan, S., com 2706 citações.
+Tal impacto deve-se, sobretudo ao artigo \emph{A Survey on Transfer Learning}, que é o artigo mais citado da área com 2240 citações.
+O mérito deste trabalho é apresentar definições, notações e uma taxonomia para TL.
+Saiu na revista \emph{IEEE Transactions on Knowledge an Data Engineering}, de 2,775 de fator de impacto, e ocupa a 33\textordfeminine ~posição no ranking de publicações InCite JCR para a categoria \emph{Computer Science, Artificial Intelligence}, ou seja, uma publicação cujo foco não é o tipo de pesquisa que menciona \emph{transferência de aprendizado}.
+
+  A China é o país com o maior número de artigos publicados, seguido pelos Estados Unidos.
+O Brasil está na 11\textordfeminine ~colocação, junto com a Índia, mas a frente de países como Espanha, Coreia do Sul e Israel.
+
+  A maior parte dos artigos são publicados em \emph{proceeding} de conferências, 63\%.
+A conferência com mais artigos sobre TL é a CVPR\footnote{Conference on Computer Vision and Pattern Recognition}, mas a ECCV\footnote{European Conference on Computer Vision}, apesar de ser somente a 4\textordfeminine em artigos,  é a conferência com o maior número agregado de citações.
+É digno de nota que as conferências de visão computacional tem sido o espaço onde artigos de TL mais são divulgados.
+Ao mesmo tempo, estranha a falta de alguma conferência com maior foco em NLP.
+Também não há entre os 20 artigos mais citados, nenhum que seja especializado em linguagem.
+  \subsection{Os Clássicos}\label{classicos}
+  \begin{figure}[h!]
+    \fbox{\includegraphics[width=\columnwidth]{completo_21.pdf}}
+    \source{Web of Science (março/2019)}{VosViewer\protect{~\cite{VOSviewer}}}
+    \caption{Núcleos de conhecimento obtidos pela análise de co-citações. Os diferentes grupos representam autores que normalmente são co-citados nos 1268 artigos resultantes da busca realizada.}
+    \label{fig:classicos}
+  \end{figure}
+  \begin{figure*}[htp]
+    \centering
+    \fbox{\includegraphics[width=\textwidth]{listas.pdf}}
+    \source{Web of Science (março/2019)}{Excel}
+    \caption{Panorama da produção científica sobre Transferência de Aprendizado.} \label{fig:toptop}
+  \end{figure*}
+A análise de co-citação usando o \emph{VosViewer} apontou para existência de 3 núcleos\footnote{\emph{clusters}} de conhecimento.
+Esses núcleos têm um forte componente temporal, podendo ser vistos como ondas: a primeira concentrando obras anteriores a 2011 (a moda é 2006),  está em amarelo na figura \ref{fig:classicos}; a segunda entre 2011 e 2014 (a moda é 2012) está em vermelho; e a terceira e última onda, em verde na figura, são obras de 2012 até hoje (moda 2014).
+
+\subsubsection{Primeira onda}\label{primeira_onda}
+Uma das características mais fortes desta onda é sua componente teórica.
+Nesse núcleo estão obras que apresentam algumas classes de transferência de aprendizado: Thurn, S.
+(1996) \cite{thrun1996learning} e Caruana, R.
+(1997) \cite{Caruana1997} introduzem \emph{Multi-Task Learning} e a ideia de que tarefas auxiliares introduzem um viés indutivo que facilita a convergência do aprendizado; Chapelle, O.
+(2006) \cite{Chapelle:2010:SL:1841234} com seu livro sobre \emph{Semi-Supervised Learning}; Raina, R.
+(2007) e \emph{Self-taught learning}; Vapnik, V.
+(1998) \cite{Vapnik1998} e os fundamentos da teoria de aprendizagem estatística.
+
+Uma outra componente da primeira onda são artigos de Adaptação de Domínio, com certo foco em abordagens que buscam aprender características latentes, de menor dimensionalidade, com o pressuposto que em um espaço latente os domínios são mais similares: Ando, R.
+K.(2005) \cite{ando2005framework}, Blitzer, J.
+(2006) \cite{Blitzer:2006:DAS:1610075.1610094} e Daume, H.
+(2006) \cite{DaumeIII2006}, por exemplo.
+
+Por fim, algumas das obras mais citadas desta onda são \emph{surveys}: Pan, S.
+(2010) \cite{PanYang}, Taylor, M.
+E.
+(2009) \cite{Taylor:2009:TLR:1577069.1755839}; que não por coincidência aparecem já mais para o final da primeira onda, organizando a pesquisa até aquele ponto.
+
+\subsubsection{Segunda onda}\label{segunda_onda}
+A segunda onda é dominada por artigos sobre Adaptação de Domínio.
+Alguns continuam com o foco em características latente como Pan, S.
+(2011) \cite{Pan2011}, mas a maioria usa abordagens que tentam selecionar amostras do domínio fonte que sejam similares ao domínio alvo: Ben-David, S.
+(2010) \cite{BenDavid2009} aborda similaridade da distribuição de domínios, Si, S.
+(2010) \cite{SiSi2010} busca minimizar a divergência; Nesta abordagem também é comum utilizar classificadores, em geral \emph{Support Vector Machines} (SVMs): Yang, J.
+(2007) \cite{Yang2007}, Bruzzone, L.
+(2010) \cite{Bruzzone2010}, Duan, L.
+(2012)\cite{LixinDuan2012}, entre outros.
+
+Há também trabalhos de Adaptação Não Supervisionada de Domínio: Gong, B. (2012) \cite{BoqingGong2012}, Fernando, B. (2013) \cite{Fernando2013}.
+Notou-se com estranheza que não apareceu em nossa análise nenhuma \emph{survey} "fechando" essa onda, apesar de sabermos da existência do capítulo \emph{A Comprehensive Survey on Domain Adaptation for Visual Applications}, Csurka, G.
+(2017) \cite{Csurka2017}.
+Foi constatado, entretanto, que esta obra não foi indexada pela base \emph{WoS} e portanto não fez parte dos resultados de nossas buscas.
+
+\subsubsection{Terceira onda}\label{terceira_onda}
+A terceira onda é o núcleo de conhecimento de transferência de aprendizado no contexto de \emph{deep learning}, que podemos chamar de \emph{deep transfer learning}.
+
+Encontramos obras clássicas como: Hinton, G.
+(2006) \cite{Hinton2006}, talvez o artigo seminal de \emph{Deep Learning}; Bengio, Y.
+(2009) \cite{Bengio2009} apresenta o conceito de \emph{representation learning} e a ideia de curriculum, uma lista de tarefas a serem aprendidas em sequencia, por terem representações em crescentes níveis de complexidade; LeCun, Bengio e Hinton (2015)\cite{LeCun2015} fazem uma revisão da área de \emph{Deep Learning}.
+Também aqui se encontra o artigo de Krizhevsky, A.
+(2012) \cite{alexnet}, sobre \emph{AlexNet}, o modelo que, ao ganhar o desafio ImageNet (ILSVRC) de 2012 com uma margem de mais de 40\% acima do segundo colocado, tornou \emph{Deep Learning} conhecido do público geral e deu início a uma verdadeira "corrida do ouro" ~na pesquisa de aprendizado de máquina.
+
+Alguns artigos são sobre \emph{large datasets}, um importante componente de \emph{deep learning}:  Deng, J.
+(2009) \cite{Deng2009} apresenta a ImageNet e sugere que modelos treinados nela poderão ser usados para aprender novos domínios mais facilmente; Everingham, M.
+(2009) \cite{Everingham2009} traz o desafio Pascal VOC; Russakovsky, O.
+(2015) \cite{Russakovsky2015} analisa o impacto da ImageNet em diversos problemas de visão computacional, e, consequentemente, o papel da transferência de aprendizado.
+
+Há também artigos que apresentam arquiteturas para resolução de tarefas específicas, utilizando modelos pre-treinados na ImageNet como extração de características: Girshick, R.
+(2014) \cite{Girshick2014}, modelo R-CNN, He, K.
+(2016) \cite{He2016}, Resnet, e Simonyan, K.\cite{simonyan2014very}, em detecção de objetos;  Fan, J.
+\cite{JialueFan2010} em rastreamento de humanos; Long e Shelhamer (2015) \cite{Long2015} em segmentação semântica; entre outros.
+
+Transferência de aprendizado passou a ser tão onipresente em \emph{Deep Learning} que Mahajan, D.
+(2018) \cite{mahajan2018exploring} argumenta que \emph{não} pré-treinar modelos na ImageNet em problemas de visão computacional é um descuido.
+
+Por essa razão, é difícil apontar trabalhos desta onda cujo foco seja apenas transferência de aprendizado.
+Nessa vertente, alguns artigos dignos de nota são: Glorot, X.
+(2011) \cite{glorot2011domain} é o primeiro a propor aplicar \emph{deep neural networks} par aprender representações comuns entre domínios; Donahue, J.
+(2014)\cite{donahue2014decaf} e Oquab, M.
+(2014)\cite{Oquab:2014:LTM:2679600.2680210} mostram que parâmetros aprendidos em tarefas fonte servem como representações que permitem aprender com poucos exemplos a tarefa alvo, dando um embasamento teórico a técnica mais largamente utilizada na prática, sintonia fina.
+\subsubsection{Resultados da análise textual}
+Os 20 artigos mais citados representam quase metade das citações (figura \ref{fig:citacoes_por_ano}).
+Assim, assumimos que uma análise textual nesse subconjunto da "Busca10Anos" é um bom \emph{proxy} do conjunto dos artigos.
+
+A figura \ref{fig:scatterText_documentbased} apresenta os 20 artigos mais citados visualizados por proximidade dos termos contidos nos \emph{abstracts} e \emph{keywords} dos mesmos.
+Os eixos são os dois termos com maior \textbf{tf-idf}.
+Quando o artigo faz parte de algum dos núcleos clássicos (figura \ref{fig:classicos}), representamos o seu ponto no gráfico com a mesma cor do núcleo clássico do qual faz parte.
+É interessante notar, que em linhas gerais, por um método totalmente diferente, agrupamos nos mesmos núcleos os artigos.
+Na figura \ref{fig:scatterText_documentbased}, cada quadrante equivale a um núcleo da figura \ref{fig:classicos}.
+Essa é uma forte validação das análises.
+
+A figura \ref{fig:clouds} apresenta "nuvens de palavras" para cada quadrante da figura \ref{fig:scatterText_documentbased}.
+É interessante notar que esses termos validam a análise feita na seção \S\ref{classicos}.
+
+\begin{figure}
+  \fbox{\includegraphics[width=\columnwidth]{top20_abstracts.pdf}}
+  \source{WoS (março/2019)}{ScatterText~\protect{\cite{kessler2017scattertext}}}
+  \caption{Os 20 artigos mais citados visualizados por proximidade de \emph{bag of words}.} \label{fig:scatterText_documentbased}
+\end{figure}
+\begin{figure}
+  \fbox{\includegraphics[width=\columnwidth]{keyword_clouds}}
+  \source{WoS (março/2019)}{TagCrowd}
+  \caption{Nuvens de palavras dos diferentes quadrantes da figura \ref{fig:scatterText_documentbased}.} \label{fig:clouds}
+\end{figure}
+
+\subsection{As Fronteiras}\label{fronteiras}
+  Pra caracterizar as fronteiras de pesquisa em transferência de aprendizado fizemos duas análises:
+  \subsubsection{Análise Textual}
+
+  Utilizamos o \emph{ScatterText}~\cite{kessler2017scattertext} para analisar as diferenças dos termos que representam os artigos da  "Busca3anos", que são mais prováveis de estar na fronteira de pesquisa, com os termos da "Busca10Anos", que engloba todos os artigos sobre TL.
+
+  O resultado é apresentado na figura \ref{fig:scatterText}. 
+  É interessante notar que os termos mais identificados com as fronteiras são \emph{deep}, \emph{neural}, \emph{images} e também palavras como \emph{cnn},  \emph{trained networks} e \emph{datasets}.
+O que tem bastante correspondência com a noss análise da terceira onda em \S\ref{terceira_onda}.
+Já os termos da "Busca10Anos" são: \emph{distribution}, \emph{domain}, \emph{adaptation}, \emph{auxiliary} (de tarefas auxiliares), \emph{kernel}; bastante similariedade com a segunda onda \S\ref{segunda_onda}.
+
+  \begin{figure}
+    \includegraphics[width=\columnwidth]{frontier.pdf}
+    \source{\emph{WoS}(março/2019)}{ScatterText\protect{\cite{kessler2017scattertext}}}
+    \caption{Análise visual dos termos mais "Fronteira" versus os termos mais "Clássico" no âmbito de Transferência de Aprendizado.} \label{fig:scatterText}
+  \end{figure}
+
+  \subsubsection{Acoplamento Bibliográfico}
+  O resultado do acoplamento bibliográfico dos artigos da "Busca3anos" pode ser analisado na figura \ref{fig:bicoupling}.
+  \begin{figure}[b]
+    \fbox{\includegraphics[width=\columnwidth]{frontier_heat}}
+    \source{\emph{WoS}(março/2019)}{VosViewer}
+    \caption{Mapa de calor da análise de acomplamento bibliográfico.} \label{fig:bicoupling}
+  \end{figure}
+
+  Analisando os artigos mais destacados, temos:
+  Lucena, O.
+(2017) \cite{Lucena2017} e de Rezende, E.
+(2017) \cite{Rezende2017}, ambos artigos brasileiros, são aplicações de \emph{deep convolutional networks} para novos problemas, \emph{face anti-spoofing} e \emph{identificação de fake images}, respectivamente.
+  Venkateswara, H. (2017) \cite{Venkateswara2017} e Ge, W. (2017)\cite{Ge2017} propõem novos métodos de Adaptação de Domínio usando \emph{Deep Learning}. 
+  Fica claro que o aspecto em comum de todos esses trabalhos é que são exemplos de \emph{Deep Transfer Learning}.
+
+  \subsubsection{Os clássicos de amanhã}
+  As análises quantitativas, apontam grande potencial em \emph{Deep Transfer Learning}, mas não são capazes de sugerir se esse ou aquele trabalho é ou não promissor.
+Para tanto, além de escolher trabalhos com as características mais "quentes", também é necessária uma análise subjetiva dos trabalhos que parecem estar avançando em espaços em aberto.
+
+  Nesse espírito, ressaltamos algumas obras que acreditamos ter potencial para se tornarem clássicos de amanhã:
+  \begin{itemize}
+    \item Zamir, A.
+(2018)\cite{taskonomy}: considerado o melhor artigo da CVPR 2018, esse artigo explora as relações entre tarefas, criando um grafo que pode ser utilizado para definir que tarefas são as melhores fontes para uma determinada tarefa alvo, entre outras questões.
+    \item Howard, H.
+(2018)\cite{ulmfit}: o modelo ULMFit avançou o estado-da-arte de tarefas de classificação de texto entre 18 e 24\% e pode se tornar o equivalente em NLP ao que Krizhevsky, A.
+(2012)\cite{alexnet} foi para visão computacional.
+    \item Zhu, J.
+(2018)\cite{CycleGan}: propõe o uso de \emph{Generative Adversarial Networks (GANs)} para transferir o domínio de uma imagem para outra de forma totalmente não supervisionada.
+Somos extremamente otimistas em relação ao potencial de GANs para transferência de aprendizado.
+    \item Ruder, S. (2019)\cite{Ruder2019Neural}: apresenta uma taxonomia mais atual para problemas de transferência de aprendizado em NLP.
+  \end{itemize}
+
+\section{Problemas em Aberto}\label{emAberto}
+A visão geral que uma revisão sistemática traz, nos permite perceber lacunas no conhecimento científico. Alguns problemas em aberto são:
+\begin{enumerate}
+  \item inexistência de métricas específicas para transferência: \emph{"models do what you inspect, not what you expect".}
+  \item a taxonomia corrente, Pan, S.
+(2009)\cite{PanYang}, dá ênfase demais a adaptação de domínio e de menos a transferência indutiva de aprendizado.
+Também precisa ser atualizada para incluir ideias mais recentes como GANs e \emph{autoencoders}.
+  \item NLP: ainda há muito pouco em transferência de aprendizado para NLP, exceção feita a Ruder, S. (2019)\cite{Ruder2019Neural}.
+  \item Teoria: na primeira onda, o papel da teoria era apontar caminhos que poderiam dar certo.
+Hoje há vários caminhos que estão dando certo na prática, mas muito pouco se sabe  o porquê de darem certo.
+É importante saber esse porquê.
+\end{enumerate}
+
+\section{Conclusão}\label{conclusao}
+Nesta revisão sistemática da literatura, vimos que as fronteiras da transferência de aprendizado estão na tendência de \emph{deep transfer learning}, um campo com muitos espaços em aberto e que engloba o uso de \emph{deep learning} para: a) resolução de novos problemas usando tarefas fonte como extração de características para solução de tarefas alvo; b) encontrar representações latentes entre domínios; c) transferir características entre domínios de forma não supervisionada; entre outros.
+Nesse percurso, nos guiamos pelo método de enfoque meta-analítico, utilizando diferentes ferramentas de análise quantitativa para validar nossas conclusões; demonstrando que é possível embasar a avaliação das fronteiras de pesquisa com dados bibliométricos.
+
+No futuro, questões importantes precisam ser adereçadas: Primeiro, a taxonomia base de Pan, S.
+(2009)\cite{PanYang} já não condiz com a atualidade, dá importância demais ao que era importante na primeira onda \S\ref{primeira_onda} e já não classifica bem os problemas da terceira onda \S\ref{terceira_onda}.
+Além disso, o método de revisão merece diversas melhorias: 1) poderia incluir outras bases como a \emph{Scopus} e \emph{Google Scholar}; 2) Podemos ainda alargar os parâmetros de pesquisa incluindo tipos de transferência como: \emph{multi-task learning}, \emph{domain adaptation} e outros nomes que caíram em desuso como \emph{learning to learn} e \emph{lifelong learning}; 3) poderia englobar a sumarização dos principais trabalhos, de preferência adotando um \emph{framework} como o dos cinco W's e um H\cite{hohman2018visual}.
+Por fim, faz-se necessário consolidar o arcabouço teórico e "abrir a caixa-preta" da transferência de aprendizado em \emph{deep learning} e avançar ainda mais nas diferentes aplicações.
 \bibliographystyle{ACM-Reference-Format}

### 2019-05-17 relatorio/paper.tex
repo reorganization

--- 2019-04-03 paper.tex
@@
 \usepackage{enumerate}
+\usepackage[most]{tcolorbox}
 
-\usepackage[most]{tcolorbox}
 \usepackage[english,ngerman,brazilian]{babel}
@@
   float=tbp,title={#2},
-  colframe=gray!75!black,olback=yellow!5!white,#1
+  colframe=gray!75!black,colback=yellow!5!white,#1
 }
@@
     \source{\emph{WoS}(março/2019)}{VosViewer}
-    \caption{Mapa de calor da análise de acomplamento bibliográfico.} \label{fig:bicoupling}
+    \caption{Mapa de calor da análise de acoplamento bibliográfico.} \label{fig:bicoupling}
   \end{figure}

### 2019-05-17 charts/chart.py

@@ -1,0 +1,6 @@
+import pandas as pd
+# import scattertext as st
+# import spacy
+# from pprint import pprint
+
+pd.read_csv('keywords.csv')
